Prolactin-mediates a lactation-induced suppression of arcuate kisspeptin neuronal activity necessary for lactational infertility in mice

Version 1

Abstract

Abstract

The specific role that prolactin plays in lactational infertility, as distinct from other suckling or metabolic cues, remains unresolved. Here, deletion of the prolactin receptor (Prlr) from forebrain neurons or arcuate kisspeptin neurons resulted in failure to maintain normal lactation-induced suppression of estrous cycles. Kisspeptin immunoreactivity and pulsatile LH secretion were increased in these mice, even in the presence of ongoing suckling stimulation and lactation. GCaMP6 fibre photometry of arcuate kisspeptin neurons revealed that the normal episodic activity of these neurons is rapidly suppressed in pregnancy and this was maintained throughout early lactation. Deletion of Prlr from arcuate kisspeptin neurons resulted in early reactivation of episodic activity of kisspeptin neurons prior to a premature return of reproductive cycles in early lactation. These observations show dynamic variation in arcuate kisspeptin neuronal activity associated with the hormonal changes of pregnancy and lactation, and provide direct evidence that prolactin action on arcuate kisspeptin neurons is necessary for suppressing fertility during lactation.

Introduction

In mammals, lactation is accompanied by a period of infertility. This adaptive change establishes appropriate birth spacing to enable maternal metabolic resources to be directed towards caring for the new-born offspring, rather than supporting another pregnancy 1. Lactational infertility is characterized by a lactation-induced suppression of pulsatile luteinizing hormone (LH) secretion, and the temporary loss of the reproductive cycle (in rodents this is exhibited as an extended period of diestrus or anestrus) 2–5. Lactation is also characterised by chronically elevated levels of the anterior pituitary hormone prolactin, which is essential for milk production and promotes adaptive changes in maternal physiology and behaviour 1, 2, 4, 5. Despite hyperprolactinaemia being a well-recognized cause of infertility, the specific role that prolactin plays in lactational infertility, as distinct from other suckling- or metabolic-related cues, is currently unclear 4, 6.

Recent in vivo studies have confirmed that kisspeptin neurons in the arcuate nucleus of the hypothalamus are responsible for the periodic release of gonadotrophin-releasing hormone (GnRH) and subsequent pulsatile luteinising hormone (LH) secretion that drives reproductive function 7–11. Studies using GCaMP6 fibre photometry in conscious mice have demonstrated that the arcuate kisspeptin neuronal population exhibits episodes of increased intracellular calcium levels coincident with, and immediately preceding, each pulse of LH secretion in intact and gonadectomised male and female mice 7, 8 9. Miniscope investigation showed that individual kisspeptin neurons within the arcuate population act in a coordinated, synchronised, and episodic manner 10, 11. Loss of pulsatile LH secretion during lactation and consequent lactational infertility may be caused by the loss of kisspeptin-mediated stimulation of GnRH secretion 12–18. Kisspeptin expression is markedly suppressed in lactation 12, 16 and when exogenously stimulated, kisspeptin neurons are unable to activate GnRH neurons during lactation, likely due to a lack of kisspeptin synthesis 13.

It is well established that hyperprolactinemia causes infertility, and thus, the elevated prolactin present in lactation seems a likely candidate to be involved in suppressing fertility during lactation. Prolactin administration acutely suppresses LH secretion 19, and chronic exposure to elevated prolactin reduces Kiss1 mRNA expression in the arcuate nucleus 17, 20, 21. In lactating mice, suppressing endogenous prolactin secretion shortens the period of infertility 22, suggesting that prolactin is important for maintaining the suppression of pulsatile LH secretion during lactation. Such a role for prolactin is controversial 4, 6, 23–27, however, with studies in a number of species suggesting that the neural stimulation of suckling may be more important in maintaining lactational infertility 28, 29. However, it has previously been difficult to disentangle the specific role of prolactin, as suckling, prolactin, and milk production are so tightly linked that manipulating one ultimately impacts the others, making it difficult to determine the contribution of any one element. Here, using a conditional deletion strategy, we have blocked prolactin action in the brain leaving suckling, lactation, and maternal behaviour intact. Using GCaMP fibre photometry techniques, we have also documented arcuate kisspeptin neuron activity across pregnancy and lactation transitions in the same mice and established that prolactin directly acts on these neurons to suppress fertility in lactation.

Materials and Methods

Animals

All experiments were performed using adult female mice on a C57BL/6J background (8-20 weeks of age). Mice were housed under controlled temperature (22 °C ± 2°C) and lighting (12-hour light/12-hour dark schedule, with lights on at 0600 hours) with ad libitum access to food and water (Teklad Global 18% Protein Rodent Diet 2918; Envigo, Huntingdon, United Kingdom). Daily body weight was recorded and vaginal cytology was used to monitor the estrous cycle stage. All experiments were carried out with approval from the University of Otago Animal Welfare and Ethics Committee.

Mice were mated with male wild-type C57BL/6J mice (presence of sperm plug = day 1 pregnancy). The first day a litter was seen was counted as day 1 of lactation and maternal mice were left undisturbed till day 3 of lactation, when vaginal monitoring would resume and litter size was normalised to 6 pups per animal, unless otherwise stated.

To monitor pulsatile secretion of LH, serial tail tip blood sampling and measurement of LH by ELISA was undertaken as reported previously 19, 30, 31. As novel exposure and restraint stress has been shown to suppress pulsatile LH secretion 32, all mice were habituated to the tail tip blood sampling procedure in a gentle restraint device (soft cardboard tube) or hand, for at least 3 weeks prior to experimentation 33. Sequential whole blood samples (4μl) were collected in 6 minute intervals for 3 hours between 0900 and 1200 hours unless otherwise stated. Samples were immediately diluted in 48ul 0.01M PBS/0.05% Tween 20, and frozen on dry ice before being stored at −20°C for subsequent LH measurement.

Effect of neuron-specific deletion of the prolactin receptor gene on the maintenance of lactational infertility

To investigate whether prolactin action in the brain is required for lactational infertility, neuron-specific Prlr knockout mice (Prlrlox/lox/Camk2aCre) and their respective Cre-negative controls (Prlrlox/lox) were generated, as previously described 34. We have previously shown that while Prlrlox/lox/Camk2aCre mice do not have a complete Prlr deletion in the forebrain, there are areas of extensive deletion (as measured by reduced prolactin-induced pSTAT5), such as the arcuate nucleus and ventromedial nucleus of the hypothalamus, and areas where Prlr is reduced by about 50% such as the medial pre-optic area 34, 35. RNAscope in-situ hybridization was done to confirm knockdown. Briefly, intact diestrous mice and 14 day OVX mice (all aged 8-16 weeks) were perfused with 2% PFA to enable visualisation of kisspeptin cell bodies in both the RP3V and ARC regions (as kisspeptin cell bodies are only visible in the RP3V of intact mice and in the ARC of OVX mice, due to estradiol regulation36). Brain sections (14μm-thick) were prepared, thaw mounted onto superfrost-plus microscope slides and then stored at −80°C. RNAscope in-situ hybridization was performed using the RNAscope 2.5 High definition Duplex Detection kit – chromogenic (Advanced Cell Diagnostics, Hayward, CA) largely in accordance with manufacturer’s instruction. The channel 1 Prlr probe was custom designed to pick up only the long form of the prolactin receptor. It was designed to transcript NM_011169.5 with a target sequence spanning nucleotides 1107-2147 (Ref: 588621; Advanced Cell Diagnostics, Hayward, CA). The channel 2 Kiss1 probe was custom designed to transcript NM_178260.3 with a target sequence spanning nucleotides 5 to 485 (Ref: 500141-C2; Advanced Cell Diagnostics, Hayward, CA). Sections were thawed at 55°C, postfixed for 3 minutes in 2% PFA, washed in 0.01M PBS for 5 minutes, and endogenous peroxidases were blocked with a hydrogen peroxidase solution for 10 minutes. Tissue was washed in distilled water (dH2O) (3x 2 minutes), then immersed in 100% ethanol briefly, air dried for 5 minutes, and a hydrophobic barrier was applied. Tissue was permeabilized with RNAscope protease plus for 30 minutes at 40°C. Sections were washed (2x 2 minutes) and were hybridized with the Prlr and Kiss1 probes (1:300 dilution, Prlr:Kiss1) or negative control probe (Cat#320751; Advanced Cell Diagnostics, Hayward, CA) at 40°C for 2 hours. Amplification (Amp 1-6) was performed in accordance with the manufacturer’s instructions. Sections were then hybridized with a Fast-RED (1:60, Fast-RED B:Fast-RED A) for 10 minutes at room temperature, before undergoing further amplification steps (Amp 7-10) in accordance to manufacturer’s instructions. The final positive hybridization was detected by incubation with the secondary detection reagents (1:50, Fast-GREEN B:Fast-GREEN) for 10 minutes at room temperature. Sections were washed, counterstained with haematoxylin (25% Gills), dried at 60°C for 20 minutes, and cover-slipped with VectaMount (Vector laboratories, H-5000) before imaging as previously described. Quantification of the proportion of kisspeptin neurons co-expressing Prlr mRNA was undertaken in FIJI software (National Institute of Health, Bethesda, Maryland, USA) following image acquisition. The total number of Kiss1-expressing cells and the total number of these that showed Prlr mRNA expression were counted. Prlrlox/lox/Camk2aCre mice showed a significant decrease in the percentage of Kiss1- expressing cells co-expressing Prlr compared to controls in both the RP3V (p = <0.0001) and arcuate nucleus (p = 0.0009) (unpaired two-tailed t tests, Supplementary Figure 1A-D). The Prlrlox/lox/Camk2aCre mice are hyperprolactinaemic due to impaired negative feedback of prolactin on hypothalamic dopamine neurons 34 and therefore show disrupted estrous cycles (showing recurrent pseudopregnancy-like cycles with long periods of diestrus of approximately 14 days between estrus stages). However, these mice are able to become pregnant and have normal pregnancies. All mice were given a 250μl subcutaneous injection of bromocriptine (5mg/kg, 5% ethanol/saline; Tocris Bioscience Cat#0427) prior to being mated. This treatment was designed to reinstate an estrous cycle in Prlrlox/lox/Camk2aCre mice. Bromocriptine is an agonist for the type 2 dopamine receptor and inhibits prolactin secretion from the pituitary gland 37, thereby terminating the pseudopregnancy-like state and bringing the mice into proestrus the following day. Following treatment, all mice were then housed with a stud male.

For Prlrlox/lox/Camk2aCre mice, estrous cycles were monitored from day 3 of lactation until the first day of diestrus following a day of estrus (proestrus and estrus had to be observed prior to transcardial perfusion on the first day of diestrus). Brains were collected following transcardial perfusion for assessment of kisspeptin immunoreactivity. For every lactating Prlrlox/lox/Camk2aCre mouse (n = 8), the brain of a Prlrlox/lox control mouse (n = 8) of the equivalent day (±1) of lactation was also collected. A group of non-lactating (NL) mice of both genotypes (n = 5-6) was also perfused for immunohistochemistry on diestrus.

To evaluate pulsatile LH secretion in early lactation (prior to the return of estrous cycles) and to determine whether progesterone played any role in regulating pulsatile LH secretion in lactation, additional groups of lactating Prlrlox/lox/Camk2aCre and Prlrlox/lox control mice were generated and treated with either the progesterone receptor antagonist, mifepristone (4mg/kg in sesame oil, s.c.; AK Scientific Inc Cat#J10622), or vehicle (n = 7-8 per group) on the morning of day 4 of lactation and on day 5 of lactation, 30 minutes prior to blood sampling that day. This dose was selected as it was found to be sufficient to cause termination of pregnancy in wild-type C57BL/6J mice (p = 0.0072, Chi-squared test, Supplementary Figure 2A; pilot study) and neither vehicle nor mifepristone treatment had an effect on litter weight gain (interaction of time x genotype & treatment p = 0.5322, two-way repeated measures ANOVA, Supplementary Figure 2B).

Measurement of LH concentrations

An established sandwich ELISA method was used to determine LH concentration in diluted whole blood samples collected from mice 30, 31. Briefly, a 96-well high plate was incubated with bovine monoclonal antibody (LHβ518b7, 1:1000 in 1xPBS; Dr. L. Sibley, UC Davis, CA, USA) for 16 h at 4°C. Following incubation of standards, controls and experimental samples for 2 hours, plates were incubated in rabbit polyclonal LH antibody (AFP240580Rb; 1:10,000; National Hormone and Pituitary Program, NIH) for 90 minutes, followed by incubation with polyclonal goat anti-rabbit IgG/HRP antibody (1:1000; DAKO Cytomation) for 90 minutes. Finally, plates were incubated in OPD (o-phenylenediamine capsules; Sigma-Aldrich Cat#P7288) for 30 minutes. A standard curve for the detection of LH concentration was generated using serial dilutions of mouse LH-reference preparation peptide (National Hormone and Pituitary Program, NIH). Luteinizing hormone levels were read using a standard absorbance plate reader (SpectraMax ABS Plus; Molecular Devices) at 490nm and 630nm wavelengths.

PULSAR Otago was used to define LH pulses 38. Parameters used; Smoothing 0.7, Peak split 2.5, Level of detection 0.04, Amplitude distance 3, Assay variability 0, 2.5, 3.3, G(1)=3.5, G(2)=2.6, G(3)=1.9, G(4)=1.5, G(6)=1.2. Mean LH levels were calculated by averaging all LH levels collected during the experiment. The assay had a sensitivity of 0.04ng/ml to 4ng/ml, with an intra-assay coefficient of variation of 4.40% and an inter-assay coefficient of variation of 8.29%. See supplementary information, figure 1, for all individual LH profiles.

Prlrlox/lox/Camk2aCre mice do not undergo the normal period of lactational infertility and the lactation-induced suppression of kisspeptin immunoreactivity is absent.

(A) Kisspeptin immunoreactivity shown in representative photomicrographs from the rostral periventricular region of the third ventricle (RP3V) non-lactating (NL; left) and lactating (right) Prlrlox/lox control and Prlrlox/lox/Camk2aCre mice (from anteroventral periventricular nucleus (AVPV) region of RP3V). (B) Representative photomicrographs showing mid arcuate nucleus (mARC) of a lactating Prlrlox/lox mouse (top) and a lactating Prlrlox/lox/Camk2aCre mouse (bottom). (C) Total kisspeptin cell number for the RP3V (NL Prlrlox/lox (n = 6) versus lactating Prlrlox/lox control (n = 8) p = 0.0100, NL Prlrlox/lox/Camk2aCre (n = 5) versus lactating Prlrlox/lox/Camk2aCre (n = 8) p = 0.6409). Two-way ANOVA followed by Tukey’s multiple comparisons test. (D) Quantification of kisspeptin fibre density in the arcuate nucleus (Fiji software, measured in percentage voxels per region of interest), showing total kisspeptin fibre density in the arcuate nucleus (lactating Prlrlox/lox control n = 8, lactating Prlrlox/lox/Camk2aCre n = 7, p = 0.0020, unpaired two-tailed t test). (E) Prlrlox/lox/Camk2aCre mice (blue, n = 8) resume estrous cycles significantly earlier (100% within 6-10 days of lactation) than Prlrlox/lox controls (grey, n = 10) (p = <0.0001, Log-rank (Mantel-Cox) test). Scale bar image and insert = 50μm. Values are shown as mean ± SEM.

Assessment of kisspeptin expression

Perfusion and fixation of tissue

Mice were anaesthetised with sodium pentobarbital (15mg/mL) and transcardially perfused with 4% paraformaldehyde. Brains were removed, postfixed in the same solution, and cryoprotected overnight in 30% sucrose before being frozen at −80°C. Two sets of 30μm thick coronal brain sections were cut using a sliding microtome, from Bregma 1.10mm to −2.80mm. Brain sections were kept in cryoprotectant solution (pH = 7.6) at −20°C until immunohistochemistry was performed.

Immunohistochemistry

Immunohistochemistry for kisspeptin in the RP3V and arcuate nucleus was performed as previously described 39. Briefly, sections were incubated in polyclonal rabbit anti-kisspeptin primary antibody (AC 566, 1:10,000; gift from A. Caraty, Institut National de la Recherche Agronomique, Paris, France) for 48 hours at 4°C. Sections were then incubated with biotinylated goat anti-rabbit IgG (1:200, Vector biolabs, Peterborough, GK) for 90 min at room temperature, followed by incubation in an avidin-biotin complex (Elite vectastain ABC kit, Vector laboratories). The bound antibody-peroxidase complex was visualised using a nickel-enhanced diaminobenzidine (DAB) reaction, to form a black cytoplasmic precipitate.

Brain sections were imaged using an Olympus BX51 light microscope and Olympus UPlanSApo 10/20x lenses. Quantification of kisspeptin neurons in the RP3V, was undertaken by manually counting all labelled neurons present in all three subdivisions, the anteroventral periventricular nucleus (AVPV), rostral preoptic periventricular nucleus (rPVpo), and caudal preoptic periventricular nucleus (cPVpo, bregma 0.02) (2 sections per brain region per mouse) and then averaging this for each animal. As kisspeptin cell bodies in the arcuate nucleus were not easily observed, as previously reported 40, kisspeptin fibre immunoreactivity was imaged using a Gryphax NAOS colour camera (Jenoptik) and evaluated using FIJI software and the voxel counter function (National Institutes of Health). Kisspeptin fibre density was measured in the arcuate nucleus across the three subdivisions; rostral arcuate (rARC), middle arcuate (mARC), and caudal arcuate (cARC) with two sections of each area per animal counted, and then averaged across each animal to get total number and reported as total amount of voxels per ROI (voxel fraction).

Characterization of arcuate kisspeptin neuronal activity using GCaMP fibre photometry

Stereotaxic surgery and AAV injections

Adult Kiss1Cre or Prlrlox/lox/Kiss1Cre mice (2-3 months old) were anaesthetised with 2% Isoflurane, given local Lidocaine (4mg/kg, s.c.) and Carprofen (5mg/kg, s.c.) and placed in a stereotaxic apparatus. A custom-made unilateral Hamilton syringe apparatus holding one Hamilton syringe was used to perform unilateral injections into the arcuate nucleus. The needles were lowered into place (−0.14mm A/P, +0.04mm M/L, −0.56mm DV) over 2 minutes and left in situ for 3 minutes before injection was made. 1μl AAV9-CAG-FLEX-GCaMP6s-WPRE-SV40 (1.3×10-13 GC/ml, University of Pennsylvania Vector Core, Philadelphia, PA, USA) was injected into the arcuate nucleus at a rate of ∼100nl/min with the needles left in situ for 3 minutes prior to being withdrawn over a period of 6 minutes. This was followed by implantation of a unilateral indwelling optical fibre (400 μm diameter, 6.5 mm long, 0.48 numerical aperture (NA), Doric Lenses, Canada, product code: MFC_400/430-0.48_6.5mm_SM3*_FLT) at the same coordinates. Carprofen (5mg/kg body weight, s.c.) was administered for post-operative pain relief. After surgery, mice received daily handling and habituation to the photometry recording procedure over 4-6 weeks before experimentation began.

GCaMP6 fibre photometry

Photometry was performed as reported previously 9. Fluorescence signals were acquired using a custom-built fibre photometry system made primarily from Doric components. Violet (405nm) and blue (490nm) fibre-coupled LEDs were sinusoidally modulated at 531 and 211 Hz, respectively, and focused into a 400μm, 0.48 numerical aperture fibre optic patch cord connected to the mouse. Emitted fluorescence was collected by the same fibre and focused onto a femtowatt photoreceiver (2151, Newport). The two GCaMP6s emission signals were collected at 10 Hz in a scheduled 5s on/15s off mode by demodulating the 405nm (non-calcium dependent) and 490nm (calcium dependent) signals. The power output at the tip of the fibre was set at 50μW. Fluorescent signals were acquired using a custom software acquisition system (Tussock Innovation, Dunedin, New Zealand) and analysed using custom templates created by Dr Joon Kim (University of Otago, Dunedin, New Zealand) based on mathematics and calculations similar to those previously described 41, 42. Briefly, the fluorescent signal obtained after stimulation with 405nm light was used to correct for movement artefacts as follows: first, the 405nm signal was filtered using a savitzky-golay filter and fitted to the 490nm signal using least linear square regression. The fitted 405nm signal was then subtracted and divided from the 490nm signal to obtain the movement and bleaching corrected signal. The output of these templates is 490-adjusted405/adjusted405, which was multiplied to get the final ΔF/F as a percentage increase (all photometry data reported as ΔF/F(%)).

All recordings were obtained from freely behaving mice for up to 24 hours and occurred between the hours of 0800 hours and 1200 hours (apart from 24 hours post weaning recording (0900 hours to 1700 hours), and day 18/19 pregnancy recording (1800 hours to 0800 hours the following day). Synchronized events (SE) were defined as when ΔF/F exceeds 3 standard deviations (SD) above the trace mean. Manual event shape analysis was performed in addition to standard deviation method for certain datasets where necessary. Events were counted manually to determine frequency of events per 60 minutes. The between animal variability in total signal means that changes in SE amplitude can only be reported as relative changes within an animal. Relative SE amplitude was calculated by using normalised ΔF/F data and then subtracting the peak of an SE from the nearest nadir to the rise of the SE and averaging that for the number of SEs in a recording. To obtain normalised ΔF/F, three pre-pregnancy datasets from each mouse were used to find the average maximum ΔF/F for that mouse. All datasets were then divided by that normalisation value to get normalised ΔF/F for each trace for each individual mouse.

Monitoring the activity of arcuate kisspeptin neurons across different reproductive stages in the same mice

Adult Kiss1Cre mice were 8-10 weeks of age at the beginning of experiments, and up to 12 months in age by time of final recording (n = 8 during pregnancy, n = 6 during lactation; 2 mice were euthanised due to dystocia therefore those mice were only followed through pregnancy). Monitoring of vaginal cytology and weights was continuous from 1 week pre-surgery till day 19 of pregnancy and resumed on day 3 of lactation (with all handling stopped on day 19 to avoid potential compromise of parturition and onset of maternal behaviour). To investigate the activity of the arcuate kisspeptin population across different reproductive states in the same animal, the following recording protocol was followed for all Kiss1Cre mice, unless otherwise stated; virgin (diestrus), day 4 of pregnancy, day 14 of pregnancy, day 18/19 of pregnancy (overnight), day 7 of lactation, day 14 of lactation, day 18 of lactation, 24 hours after weaning, first diestrus after estrous cycles begin following weaning, and 10 days after OVX. In addition, blood sample collection for paired LH measurement was done in virgin (diestrus) state, on day 14 of pregnancy (in 4 mice, maximum of 6 samples were collected around small “peaks” in baseline), day 7 of lactation, and day 14 of lactation. Blood sampling was not carried out at additional time points as blood sampling was undertaken at least a week apart, and stress from repeated sampling was attempted to be kept at a minimum e.g., not blood sampling around the time of birth. Fibre photometry recordings were usually between 2-4 hours in length. The only longer recordings were undertaken on day 18/19 of pregnancy (14 hours) and 24 hours after weaning (8 hours). These particular recording sessions were extended to determine whether there were any longer-term changes occurring in the activity of the arcuate kisspeptin population in the lead up to parturition or following weaning of pups (states closely followed by postpartum estrus and resumption of normal estrous cycles, respectively).

Effect of arcuate kisspeptin neuron-specific deletion of the prolactin receptor gene on the maintenance of lactational infertility and the activity of kisspeptin neurons during lactation

Kisspeptin-specific prolactin-receptor knockout mice (Prlrlox/lox/Kiss1Cre 19) and their respective Cre-negative controls (Prlrlox/lox) were generated. RNAscope in-situ hybridization was done to confirm knockdown, with Prlrlox/lox/Kiss1Cre mice showing a significant decrease in the percentage of Kiss1-expressing cells co-expressing Prlr compared to controls in the arcuate (p = <0.0001, unpaired two-tailed t test, Extended data Figure 3E, F). Similar to experiments described above, Prlrlox/lox/Kiss1Cre (n = 27) and Prlrlox/lox control (n = 30) dams underwent estrous cycle monitoring from day 3 of lactation onwards to determine whether mice showed an early resumption of estrus cycles.

To determine whether the deletion of the prolactin receptor from arcuate kisspeptin neurons led to early reactivation of these neurons during lactation, adult Prlrlox/lox/Kiss1Cre mice (n = 5) and an additional Kiss1Cre control mouse (n = 1) (8 weeks old at the start of the experiment, and up to 14 months at end of final recording timepoint) were set up for fibre photometry, as described above. Monitoring of vaginal cytology and weights was continuous from 1 week pre-surgery till day 18 of pregnancy and resumed on day 3 of lactation. Recordings were undertaken in a similar timeline as described above, however no pregnancy recordings were done, and in early lactation recordings were performed every 2 days from day 3 to day 9 of lactation, before following the same protocol as described. No blood samples were taken over this lactation period in this genotype. As described previously, recordings were kept between 2-4 hours, apart from the 24 hours after weaning recording (8 hours).

Statistical analysis

Data are presented as mean ± SEM and all statistical analysis was performed with PRISM software 10 (GraphPad Software, San Diego, CA, USA) with a p value of < 0.05 considered as statistically significant. Individual symbols in graphs represent individual mice. Differences in kisspeptin cells number or fibre density was assessed using two-way ANOVAs with Tukey’s multiple comparisons tests or t tests, with both analyses using combined averages of each animal (averaged number of cells or fibre density across the three subdivisions of each nucleus to get total number reported). Resumption of estrous cycles was analysed using Log-rank (Mantel-Cox) test chi square test. LH pulse frequency data and mean LH data was analysed using two-way ANOVAs with Tukey’s multiple comparisons tests. SE frequency and amplitude throughout reproductive cycles was analysed using mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests where appropriate and day 18/19 of pregnancy data was analysed using t tests. Correlation between SE occurrence and LH pulses was assessed using chi-square test. All fibre photometry data used for quantitative analysis and comparison were from the first pregnancy and lactation. A full list of probability values, inferential statistics, and degrees of freedom for all data can be found in Supplementary Table 1.

Results

Prolactin action on forebrain neurons is necessary to maintain lactational infertility

Lactation has previously been associated with a marked decrease in Kiss1 mRNA levels in both rostral periventricular region of the third ventricle (RP3V) and arcuate nucleus populations during lactation 13. To determine whether prolactin is involved in the maintenance of lactational anestrus, the Prlr gene was knocked out of Camk2a expressing neurons (most forebrain neurons, as described in 34) of female mice. Control Prlrlox/lox mice showed a marked reduction in kisspeptin cell body immunoreactivity in the RP3V of lactating compared to virgin mice (p = 0.0100, Post hoc Tukey’s multiple comparisons test, Figure 1A, C). In contrast, lactation-induced suppression of kisspeptin cell bodies in the RP3V and fibre labelling in the arcuate nucleus was absent in Prlrlox/lox/Camk2aCre mice (p = 0.6409, Post hoc Tukey’s multiple comparisons test, Figure 1A, C; interaction between reproductive state and genotype p = 0.0034, two-way ANOVA, Figure 1A, C; p = 0.0020, unpaired two tailed t test, measured as percentage voxels within the region of interest, Figure 1B, D).

Estrous cycles during lactation were significantly altered by deletion of Prlr in the forebrain, with all Prlrlox/lox/Camk2aCre mice showing a return to estrus between day 6 and day 10 of lactation (Figure 1E), while, as normal, estrus did not occur until day 20 in control Prlrlox/lox mice (p = <0.0001, log-rank (Mantel-Cox) test, Figure 1E). No differences in litter weight gain from day 3 to day 8 of lactation were observed in either group (p = 0.3282, mixed analysis test, Supplementary Figure 3), indicating that the suckling stimulus that mice received was maintained in the absence of Prlr expression in Camk2a expressing neurons. Collectively, these data show that prolactin action in the brain is absolutely required for the lactation-induced suppression of kisspeptin expression and to maintain lactational infertility in mice.

In a separate cohort of mice, pulsatile LH secretion in Prlrlox/lox/Camk2aCre mice was monitored in early lactation, prior to the return of estrous cycles. To rule out a potential role for progesterone in suppressing fertility during lactation 9, 43, 44, the progesterone receptor antagonist mifepristone (RU486), was administered to mice in early lactation. Vehicle-treated Prlrlox/lox control mice showed the expected near complete absence of pulsatile LH secretion during lactation (Figure 2A, B). In contrast, nearly all Prlrlox/lox/Camk2aCre mice showed a lack of the normal lactation-induced suppression of pulsatile LH secretion demonstrated by a significant increase in frequency of LH pulses compared to controls (effect of genotype p = 0.0024, two-way ANOVA with Tukey’s multiple comparisons test, Figure 2A, B). There was no effect of mifepristone on this pattern of LH secretion, suggesting that progesterone action is not required for the suppression of LH secretion (LH pulse frequency (interaction genotype x treatment p = 0.2807; treatment p = 0.8588; Figure 2B), mean LH levels (interaction genotype x treatment p = 0.8697; treatment p = 0.8586; Figure 2C), two-way ANOVA with Tukey’s multiple comparisons test). These data indicate that prolactin is the primary signal responsible for the suppression of LH during lactation in mice. Individual LH profiles from all animals are shown in Supplementary Figure 4.

Prolactin action in the brain during lactation is necessary for the suppression of pulsatile LH secretion.

Examples of pulsatile LH levels in the blood from lactating Prlrlox/lox controls and lactating Prlrlox/lox/Camk2aCre mice that have either been treated with vehicle (sesame oil, s.c., grey, veh) or 4mg/kg mifepristone (in sesame oil, s.c., blue, mif) on the day prior and on the day of blood sampling (2 injections). Asterisks indicate LH pulse peaks as detected by PULSAR Otago analysis. Graphs show LH pulse frequency (B; interaction p = 0.2807, genotype p = 0.0024, state p = 0.8558), and mean LH levels (C; interaction p = 0.8697, genotype p = 0.0031, state p = 0.8586). Lactating vehicle-treated Prlrlox/lox (n = 8), lactating mifepristone-treated Prlrlox/lox (n = 8), lactating vehicle-treated Prlrlox/lox/Camk2aCre (n = 8), lactating mifepristone-treated Prlrlox/lox/Camk2aCre (n = 7). Two-way ANOVA followed by Tukey’s multiple comparisons test. Values are shown as mean ± SEM.

Episodic activity of arcuate kisspeptin neurons is suppressed during pregnancy and most of lactation

To directly assess the role of prolactin in regulating kisspeptin neuron activity during lactation, GCaMP6 fibre photometry was undertaken to monitor real-time activity of arcuate kisspeptin neurons in freely behaving mice. We first undertook a longitudinal assessment of changes in kisspeptin neuronal activity by tracking individual animals throughout pregnancy and lactation and following weaning (Figure 3).

Arcuate kisspeptin neuron GCaMP6 population activity throughout different reproductive states in the same mice.

Representative neuronal activity from three Kiss1Cre mice throughout the virgin, pregnant, lactating, and post-weaning states. The time points monitored in order were: virgin diestrus, day 4 pregnancy, day 14 pregnancy, day 18/19 pregnancy (overnight), day 7 lactation, day 14 lactation, day 18 lactation, 24 hours after weaning (day 22 postpartum), return to normal cycling following weaning (return to estrous cycles), and 10 days post ovariectomy (OVX). Asterisks indicate SEs. Note: dataset from (B) on day 4 of pregnancy onwards and OVX datasets from all mice are on a different y axes scale.

Initially, GCaMP fibre photometry recordings were collected in the virgin diestrous state, both with and without serial blood sampling to measure LH concentrations. As can be seen in Figure 4A, photometry recordings were characterized by discrete synchronised events (SEs) of elevated intracellular calcium (indicative of synchronous activity of the kisspeptin population), with each SE correlating perfectly to a single pulse of LH in the minutes following (p = <0.0001, chi-squared test). These were observed to be at a similar rate to that described previously in diestrous mice using GCaMP photometry in a different Kiss1 mouse line 9, with periodic SEs occurring about once per hour (1.250±0.250; Figure 3 & 4B).

Synchronised Ca2+ events are perfectly correlated to pulsatile LH secretion across different reproductive states.

(A) When fibre photometry was paired with serial blood sampling for pulsatile LH secretion, the relationship between SEs and LH pulses was examined. Each of the times an SE was seen during a recording with blood sampling, a pulse of LH was also observed, with 100% correlation (p = <0.0001, Chi-squared test; 73 out of 73 SEs observed lead to an LH pulse). Representative examples of paired photometry and blood sampling are shown from the diestrous state (virgin), from day 7 lactation where no SEs corresponds with no LH release, and from day 14 lactation when SE are beginning to re-emerge. (B) Quantitative analysis of SE frequency per hour across different reproductive states in Kiss1Cre mice (p = 0.0012, mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests). (C) Quantitative analysis of SE amplitude of normalised ΔF/F across different reproductive states (p = 0.0118, mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests). Black asterisks indicate SEs, red asterisks indicate LH pulse peaks as detected by PULSAR Otago analysis. Values shown as mean ± SEM.

The activity of the arcuate kisspeptin population in Kiss1Cre mice dynamically changed depending on the reproductive state of the mouse (p = 0.0012, mixed-effect analysis, Figure 3 & 4B). On day 4 of pregnancy, SE frequency had markedly decreased (0.297±0.136/hr; Figure 3 & 4B), indicating an early reduction in activity of arcuate kisspeptin neurons during pregnancy. By day 14 of pregnancy, no SEs were seen (0±0/hr; Figure 3 & 4B) and this was confirmed by a lack of pulsatile LH secretion (Supplementary Figure 5). In late pregnancy (day 18), neuronal activity was monitored for 14 hours, during which time low amplitude, SEs were unexpectedly observed at similar frequencies to virgin levels (2.043±0.940/hr; Figure 3 & 4B-C). This unusual pattern of activity is illustrated in more detail in Figure 5 where alongside the resurgence of low amplitude SEs, there was a marked increase in baseline activity observed, relative to other stages. This activity was reminiscent of the miniature SEs observed to be caused by activation of subgroups of cells in a brain slice technique 11, but we are unable to resolve such events using the present methods. Since our aim was to continue longitudinal assessment of the arcuate kisspeptin population into lactation and weaning, mice were not disrupted by blood sampling immediately prior to parturition as we were concerned that this additional stressor might interfere with establishment of maternal behaviour. Hence, we are unable to report whether these low amplitude SEs and elevated baseline activity were associated with LH secretion.

Activity of arcuate kisspeptin neurons on day 18/19 of pregnancy

Fibre photometry recordings of mice on the evening of day 18 of pregnancy (0600 hours) to the morning of day 19 of pregnancy day 19 (0800 hours) shows low amplitude SEs. 3-hour section blown up for ease of viewing. (C) No difference is seen between frequency of SEs (per 60 minutes) in the virgin diestrus versus D18/19 pregnancy (p = 0.4063, paired two-tailed t test), however a significant decrease in relative SE amplitude is seen (D; p = 0.0141, paired two-tailed t test). Asterisks indicate SEs. Dotted line in insert of (A) and (B) indicates 3 standard deviations (3SD). Grey shaded region = lights off.

Evaluation of activity of arcuate kisspeptin neurons during lactation showed complete suppression of activity on day 7 of lactation with a corresponding absence of pulsatile LH secretion (0±0/hr; Figure 3 & 4B). This lactation-induced suppression of activity was partially relieved by day 14 of lactation (0.583±0.083/hr; Figure 3 & 4B), with SEs again corresponding to low frequency pulses of LH secretion. Further increases in SE frequency were seen on day 18 of lactation (0.850±0.100/hr; Figure 3 & 4B), including an increase in baseline activity, similar to that seen in late pregnancy, and by 24 hours after weaning (day 22 postpartum) the frequency of SEs returned to close to non-pregnant levels (1.375±0.114/hr; Figure 3 & 4B). Frequency remained unchanged on the day of first diestrus following a return to estrous cycles after weaning (1.533±0.226/hr; Figure 3 & 4B). As a final manipulation, mice were ovariectomized (OVX), and arcuate kisspeptin population activity observed to increase significantly with clusters of high amplitude activity (4.778±0.222/hr; Figure 3 & 4B), consistent with previous reports following OVX in nulliparous mice 45. Collectively, these observations show extensive, dynamic variation in activity of the arcuate kisspeptin neuronal population associated with pregnancy and lactation.

Mice with an arcuate kisspeptin neuron-specific deletion have premature reactivation of estrous cycles and neuronal activity in lactation

To determine whether the prolactin-induced suppression of estrous cycles and LH pulsatile secretion was specifically mediated by kisspeptin neurons, mice were generated with an arcuate-specific deletion of the Prlr from kisspeptin neurons 19. Similar to the data from forebrain neuron-specific deletion of Prlr, there was early resumption of estrous cycles in Prlrlox/lox/Kiss1Cre mice during lactation (63% showing estrus by day 10 of lactation and 83% by day 19 lactation) compared to Prlrlox/lox controls (4% by day 19 lactation) (p = <0.0001, log-rank (Mantel-Cox) test, Figure 6A). No difference in litter weight gain during lactation (day 3-20 of lactation weight gain) was observed in either group (p = 0.6404, two-way ANOVA, Supplementary Figure 1H) indicating that suckling and/or lactation itself was not impaired. In vivo GCaMP6 fibre photometry in Prlrlox/lox/Kiss1Cre mice showed early reactivation of the arcuate kisspeptin population between day 3 and 5 of lactation (Figure 6C). This was accompanied by a clear return to estrous within this early lactation window in 4/5 mice. These data demonstrate that prolactin action specifically on arcuate kisspeptin neurons is responsible for maintaining suppression of those neurons, and thereby fertility, during lactation in mice.

Prlrlox/lox/Kiss1Cre mice do not undergo the normal period of lactational infertility and show early reactivation of arcuate kisspeptin neurons prior to estrus in lactation.

(A) Prlrlox/lox/Kiss1Cre mice resume estrous cycles significantly earlier (78% within 4-18 days of lactation, n = 32) than Prlrlox/lox controls (0% by day 18, n = 30) (p = <0.0001, Log-rank (Mantel-Cox) test). (B) Representative fibre photometry traces from day 5 of lactation from either a Kiss1Cre control mouse or Prlrlox/lox/Kiss1Cre mice. Mice with Prlr knocked out of arcuate kisspeptin neurons (Prlrlox/lox/Kiss1Cre) show SEs early in lactation, which were not seen until day 14 lactation in Kiss1Cre control mice. In comparison, the Kiss1Cre control mouse shows no SEs, as seen in earlier groups sampled on day 7. Asterisks indicate SEs.

Discussion

We demonstrate here that prolactin action in arcuate kisspeptin neurons is necessary for the maintained suppression of fertility during lactation in mice. Neuron-specific Prlr deletion (Prlrlox/lox/Camk2aCre) resulted in premature return to estrus in early lactation, even in the presence of ongoing suckling stimulus and the full metabolic consequences of milk production. Accompanying the resumption of estrus was an absence of the normal lactation-induced reduction in kisspeptin immunoreactivity 12, 16, 20, and pulsatile LH secretion was also observed on day 5 of lactation prior to the premature estrus when it would normally have been completely absent 46–48. To evaluate the specific role of kisspeptin neurons in mediating the prolactin-induced suppression of fertility, we have comprehensively mapped the activity of arcuate kisspeptin neurons throughout a full reproductive cycle: pregnancy, lactation, and after weaning in individual animals. The data show an immediate suppression of activity of arcuate kisspeptin neuronal activity during pregnancy, and this is maintained throughout most of lactation, apart from a brief window of reactivation immediately prior to parturition. Deleting Prlr specifically from arcuate kisspeptin neurons prevented the suppression of activity in early lactation, resulting in premature induction of episodic activation of kisspeptin neurons, and early onset of estrus. Combined, these data provide direct evidence that prolactin action on kisspeptin neurons is necessary for lactation-induced infertility in mice.

It is now well established that the arcuate kisspeptin neurons form the GnRH “pulse generator”, and hence drive pulsatile release of GnRH from the hypothalamus and consequent pulses of LH form the pituitary that is required for fertility 49–51. This is the first study to monitor activity of the GnRH “pulse generator” across different reproductive states in the same animal, and the data largely match previously described patterns of LH secretion 47, 52. The frequency and dynamics of the synchronised events changed dramatically, initially due to the pregnancy-induced changes in ovarian hormones. The abrupt decrease in “pulse generator” activity in early gestation is likely caused by rising levels of progesterone, known to profoundly suppress activity of arcuate kisspeptin neurons and LH secretion 9, 43, 44. Progesterone is elevated throughout pregnancy, gradually increasing until luteolysis and progesterone withdrawal occurs in the lead up to parturition 53–56. Interestingly, we observed a transient reactivation of the arcuate kisspeptin neurons in the night between days 18 and 19 of pregnancy. This was characterised by frequent, low amplitude episodes of activity, and increased baseline activity that may represent the intermittent synchronized activity of small subsets of arcuate kisspeptin neurons that have not yet transitioned to full synchronization of the whole population 11. It seems likely that this pattern of activity is associated with progesterone withdrawal in late pregnancy and may be important in stimulating follicular growth leading up to a postpartum ovulation 57, 58.

In early lactation, episodic activity of arcuate kisspeptin neurons was absent, with sporadic low-amplitude activity returning around day 14 of lactation. There was another period of increased baseline activity in late lactation, similar to that seen in late pregnancy, potentially representing a signature of reactivation of synchronized activity of the arcuate kisspeptin neurons. Overall patterns of activity rapidly returned to normal diestrous levels soon after weaning. This increase in “pulse generator” activity during late lactation mirrors the increase in LH levels that has been reported as lactation progresses 59. In the absence of Prlr in arcuate kisspeptin neurons, however, synchronized episodic activity re-appeared as early as 3 days after birth, even in the presence of ongoing suckling. These data clearly show that prolactin action in the arcuate kisspeptin neurons is necessary to sustain lactational infertility in mice. The observed disruption of lactational infertility in Prlrlox/lox/Kiss1Cre mice is particularly remarkable given that Prlr deletion is restricted to arcuate kisspeptin neurons in this model 19, and prolactin action on RP3V kisspeptin neurons 21, 60 and on gonadotrophs in the pituitary gland 61–63 are unaffected.

The indispensable role for prolactin in mediating lactation-induced infertility in the mouse is surprising, given the consensus of much work in other species concluding that other factors may be more important (see 4, 6, 26, 27, 64). This may reflect a level of redundancy amongst contributing factors across all species, including ovarian hormones, metabolic cues and neural inputs of suckling. Notably, the conditional deletion approach described here distinguishes prolactin action from the neurogenic effects of suckling without altering the process of lactation itself. Moreover, this approach avoids the potential confounding effects of using dopamine agonists to suppress prolactin 28, 29, given that dopamine can directly inhibit GnRH neuronal activity 65.

While the effects of widespread neuronal deletion (Prlrlox/lox/Camk2aCre) on fertility were largely recapitulated by the arcuate kisspeptin-specific model, it was apparent that the global deletion was more effective at inducing the return to estrus during lactation (in 100% of animals by day 10), compared to the arcuate kisspeptin-specific model (63% by day 10, and 83% by day 19). This may be due to the absence of lactation-induced suppression of Kiss1 expression in RP3V kisspeptin neurons of Prlrlox/lox/Camk2aCre mice. Similarly, we cannot rule out the possibility that other populations of prolactin-sensitive neurons, such as GABA or dopamine neurons 60, may contribute to suppressing estrous cycles during lactation. Nevertheless, our data collectively provide strong evidence that prolactin action on arcuate kisspeptin neurons is the primary factor mediating lactation-induced infertility in mice. Given that hyperprolactinemia induces infertility in humans and many other species 21, 22, 66–75, it is likely that a conserved mechanism will be contributing to lactational infertility in all mammalian species.

eLife Assessment

This study addresses an important long-standing controversy in the field of reproductive biology. Using cutting-edge techniques the authors provide compelling evidence supporting a pivotal role of prolactin as the mediator of lactational infertility. Some methodological, technical, interpretive, stylistic, and typographical aspects of the paper need to be strengthened.

Reviewer #1 (Public Review):

Summary:

In this paper, Hackwell and colleagues performed technically impressive, long-term, GCaMP fiber photometry recordings from Kiss1 neurons in the arcuate nucleus of mice during multiple reproductive states. The data show an immediate suppression of activity of arc Kiss1 neuronal activity during pregnancy that is maintained during lactation. In the absence of any apparent change in suckling stimulus or milk production, mice lacking prolactin receptors in arcuate Kiss1 neurons regained Kiss1 episodic activity and estrous cyclicity faster than control mice, demonstrating that direct prolactin action on Kiss1 neurons is at least partially responsible for suppressing fertility in this species. The effect of loss of prolactin receptors from CamK2a expressing neurons was even greater, indicating either that prolactin sensitivity in Kiss1 neurons of the RP3V contributes to lactational infertility or that other prolactin-sensitive neurons are involved. These data demonstrate the important role of prolactin in suppressing Kiss1 neuron activity and thereby fertility during the lactational period in the mouse.

Strengths:

This is the first study to monitor the activity of the GnRH pulse-generating system across different reproductive states in the same animal. Another strength in the study design is that it isolated the effects of prolactin by maintaining normal lactation and suckling (assessed indirectly using pup growth curves). The study also offers insight into the phenomenon of postpartum ovulation in mice. The results showed a brief reactivation of arcuate Kiss1 activity immediately prior to parturition, attributed to falling progesterone levels at the end of pregnancy. This hypothesis will be of interest to the field and is likely to inspire testing in future studies. With the exceptions mentioned below, the conclusions of the paper are well supported by the data, and the aims of the study were achieved. This paper is likely to raise the standard for technical expectations in the field and spark new interest in the direct impact of prolactin on Kiss1 neurons during lactation in other species.

Weaknesses:

A weakness in the approach is the use of genetic models that do not offer complete deletion of the prolactin receptor from targeted neuronal populations. A substantial proportion of Kiss1 neurons in both models retain the receptor. As a result, it is not clear whether the partial maintenance of cyclicity during lactation in the genetic models is due to incomplete deletion or to the involvement of other factors. This weakness should be more fully discussed in the text. In addition, results showing no impact of progesterone on LH secretion during lactation are surprising, given the effectiveness of progesterone-containing birth control in lactating women. The progesterone-related experiments were not well justified or discussed in the text. While the authors assert their findings may reflect an important role for prolactin in lactational infertility in other mammalian species, that remains to be seen. Hyperprolactinemia is known to suppress GnRH release, but its importance in the suppression of cyclicity during lactation is controversial. Indeed, in several species, the stimulus of suckling is considered to be the main driver of lactational fertility suppression. Data from rats shows that exogenous prolactin was unable to suppress LH release in dams deprived of their pups shortly after birth; both suckling and prolactin were necessary to suppress a post ovariectomy rise in LH levels. The duration of amenorrhea does not correlate with average prolactin levels in humans, and suckling but not prolactin was required to suppress the postpartum rise in LH in the rhesus monkey. The authors should discuss more thoroughly whether the protocol of this or other studies might result in discordant results and whether mice are likely to be an outlier in their mechanism of cycle suppression.

Reviewer #2 (Public Review):

Summary:

The overall goal of Eleni et al. is to determine if the suppression of LH pulses during lactation is mediated by prolactin signaling at kisspeptin neurons. To address this, the authors used GCaMP fiber photometry and serial blood sampling to reveal that in vivo episodic arcuate kisspeptin neuron activity and LH pulses are suppressed throughout pregnancy and lactation. The authors further utilized knockout models to demonstrate that the loss of prolactin receptor signaling at kisspeptin cells prevents the suppression of kisspeptin function and results in early reestablishment of fertility during lactation. The work demonstrates exemplary design and technique, and the outcomes of these experiments are sophistically discussed.

Strengths:

This manuscript demonstrates exemplary skill with powerful techniques and reveals a key role for arcuate kisspeptin neurons in maintaining lactation-induced infertility in mice. In a difficult feat, the authors used fiber photometry to map the activity of arcuate kisspeptin cells into lactation and weaning without disrupting parturition, lactation, or maternal behavior. The authors used a knockout approach to identify if prolactin inhibition of fertility is mediated by direct signaling at arcuate kisspeptin cells. Although the model does not perfectly eliminate prolactin receptor expression in all kisspeptin neurons, results from the achieved knockdown support the conclusion that prolactin signaling at kisspeptin neurons is required to maintain lactational infertility. The methods were advanced and appropriate for the aims, the studies were rigorously conducted, and the conclusions were thoughtfully discussed. Overall, the aims of this study were achieved.

Reviewer #3 (Public Review):

Summary:

Grattan and colleagues were trying to establish the neural mechanism underlying lactational infertility, in particular trying to establish the relative importance of the neurogenic effects of the suckling stimulus versus prolactin per se. They have shown that in the mouse it is rather prolactin and more specifically its action on the hypothalamic arcuate kisspeptin neuronal system, which is the key neural construct underlying gonadotrophin-releasing hormone (GnRH) pulse generation and central to the neuroendocrine control of reproduction, that mediates lactational infertility. The authors have taken a measured tone to emphasise the data pertaining to the mouse without extravagant extrapolation to humans. Nevertheless, the key findings provide a substantial foundation to facilitate interpretation of studies in other species.

Strengths:

The major strength of this study is the use of a combination of cutting-edge technologies, which of course underlie the majority of scientific advances rather than intellectual prowess favoured by the majority of scientists. Their approach avoided the major confounding effects of using pharmacological strategies to suppress prolactin action that has complicated the vast majority of previous studies. The study also provides an elegant and comprehensive contiguous description of GnRH pulse generator frequency across the ovarian cycle, through pregnancy and lactation, and into weaning in individual animals.

Weaknesses:

There are no significant weaknesses.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

I recommend being explicit regarding how the animals were habituated to blood sampling.

On lines 109-111 we have added a more detailed explanation of how mice were habituated to blood sampling. This includes details that mice were held and had their tails palpated for approximately 5 minutes per day.

Were any mice excluded due to loss or movement of the implant over time? Any details to allow replication of long-term measurements like this should be included.

No mice lost their cannulas during experimentation so we have added a sentence on this on lines 303 to 304 to this effect. We have also noted that there was a slight decrease in signal over the months of experimentation. A statement on line 318 has also been added that clarifies two mice lost between the pregnancy and lactation stages of experiment were euthanised due to dystocia.

The text states that synchronized episodic activity reappeared as early as 3 days after birth, citing Figure 6c as evidence. There is no 6c. Figure 6b shows day 5 after birth.

This has been corrected.

The methods state mRNA levels had to be "above background" to be counted as colocalization. At how many fold/what percent above background was a cell considered positive for expression?

Positive hybridisation was scored according to the manufacturer’s protocol and a statement to this effect has been added on line 144.

Please ensure figure titles or the data graphs explicitly give the genotype of the mice in all figures (or state the mice are wildtype).

Genotype has been added to figure titles where possible. Genotypes are always given in figure legends and tables and/or explicitly stated on the figure itself.

Figure 4's title states events are "perfectly" correlated, which is a subjective term. I recommend saying "consistently" or "temporally" correlated, depending on your meaning.

This has been amended to read “consistently correlated”

Reviewer #2 (Recommendations For The Authors):

The comments below aim to clarify the paper's methodology and results but do not detract from my overall enthusiasm for this work.

- Given past studies demonstrating prolactin action in the brain, particularly the MPOA/MPN, is essential for maternal behavior, can the authors please clarify why this behavior is retained in the cam2a prlr knockout mice? The authors mention that prlr in the MPOA is only knocked down 50% compared to WT controls. Is this sufficient to retain maternal behavior?

In our experience 50% Prlr in the MPOA is sufficient to retain normal maternal behaviour in most animals including the ones in this experiment (our original paper describing this showed relatively normal behavior, for example, with a vGAT and vGlut-mediated knockouts, and even a double knockout – it was only when we achieved complete KO with an AAV-Cre that we saw failure of maternal behavior – Brown et al, PNAS 3;114(40):10779-10784 2017). We have added a statement on lines 157-159 regarding this. We have an additional paper in preparation specifically characterising the maternal behaviour and lactation outcomes in this line of mice, and we find most animals display normal maternal behaviour, with slightly impaired milk production in later lactation.

- Supplementary Figure 1. Can the authors please clarify the criteria for a cell to be positive for prlr? The methods state that the signal must be "above background level." How was the background measurement obtained? In the negative control?

As per above, scoring of positive hydribisation was done according to the manufacturer’s protocol and a statement to this effect has been added on line 144.

- Lines 310-314: This sentence describes RNAscope analysis of prlr knockdown in kisspeptin cells and refers to Extended Figure 3 - but I believe this is in Supplementary Figure 1.

This has been corrected.

- Figure 3-4: When mice return to estrous cycles, the amplitude of episodic kisspeptin neuron activity is the same as 24 hours after weaning, which appears much lower than in virgin females. Does this reach significance? If so, do the authors know why kisspeptin activity is still suppressed, and can they comment on why this may not affect estrous cyclicity?

This does not reach significance – see Supplementary Table 1 (4C) for statistics. Therefore, no further analysis was done. This question would need to be examined with a follow up experiment. Given the 5s on, 15 s off scheduled mode of recording used here, amplitude was not an extremely accurate measure and amplitude has been reported as relative within each mouse. There is also an additional issue of a gradual reduction in amplitude of signal over time in these long-term experiments – although it is true that much larger signals were detected after ovariectomy at the end of the experiment. At present, we have not tried to interpret whether the changes in amplitude are informative.

- Fiber photometry studies: Please indicate whether a post-mortem examination of GCaMP transfection and fiber photometry placement was conducted, and what region of the ARC was imaged.

Brains from these mice were collected, however postmortem analysis of cannula placement of GCaMP6 transfection was not carried out in all mice. This was based on our experience with this method, in that the quality and characteristic pattern of activity seen, as well as corresponding LH secretion following an SE, was indicative of successful cannula placement and transfection. Incorrectly placed cannular failed to show SEs. A trial was done with 3 mice and cannula placement was found to be in the caudal ARC (cARC) with GFP (attached to GCaMP) restricted to the cARC. A statement has been added on lines 306-313 regarding this.

- Were male mice removed before birth? Please add to the methods section if not included.

Yes, male mice were removed after a sperm plug was seen and were never present at parturition. We have inserted additional details on line 95 to this effect.

Reviewer #3 (Recommendations For The Authors):

(1) Line 172: n=7-8 per group, yet in Supplementary Figure 2, n=6 per group.

These are referring to different groups of mice. N=7-8 is referring to the group size of mice in Figure 2 that were given mifepristone or vehicle control. In contrast the Supplementary figure 2 n number refers to the mice in the pilot study. Additional n number for the pilot study has been added on line 194.

(2) Line 314: Extended = suppl; Figure 3 = 1.

This has been corrected.

(3) Line 451: Figure 6C, does not exist.

This has been corrected.

Line 590: Reference 23 could be replaced by Ordog T et al 1998 Am J Physiol 274,E665 because it is later and more relevant to the topic.

This reference has been replaced with the suggested reference.

Version 2

Abstract

Abstract

The specific role that prolactin plays in lactational infertility, as distinct from other suckling or metabolic cues, remains unresolved. Here, deletion of the prolactin receptor (Prlr) from forebrain neurons or arcuate kisspeptin neurons resulted in failure to maintain normal lactation-induced suppression of estrous cycles. Kisspeptin immunoreactivity and pulsatile LH secretion were increased in these mice, even in the presence of ongoing suckling stimulation and lactation. GCaMP fibre photometry of arcuate kisspeptin neurons revealed that the normal episodic activity of these neurons is rapidly suppressed in pregnancy and this was maintained throughout early lactation. Deletion of Prlr from arcuate kisspeptin neurons resulted in early reactivation of episodic activity of kisspeptin neurons prior to a premature return of reproductive cycles in early lactation. These observations show dynamic variation in arcuate kisspeptin neuronal activity associated with the hormonal changes of pregnancy and lactation, and provide direct evidence that prolactin action on arcuate kisspeptin neurons is necessary for suppressing fertility during lactation in mice.

Introduction

In mammals, lactation is accompanied by a period of infertility. This adaptive change establishes appropriate birth spacing to enable maternal metabolic resources to be directed towards caring for the new-born offspring, rather than supporting another pregnancy (1). Lactational infertility is characterized by a lactation-induced suppression of pulsatile luteinizing hormone (LH) secretion, and the temporary loss of the reproductive cycle (in rodents this is exhibited as an extended period of diestrus or anestrus) (2–5). Lactation is also characterised by chronically elevated levels of the anterior pituitary hormone prolactin, which is essential for milk production and promotes adaptive changes in maternal physiology and behaviour (1,2,4,5). Despite hyperprolactinaemia being a well-recognized cause of infertility, the specific role that prolactin plays in lactational infertility, as distinct from other suckling- or metabolic-related cues, is currently unclear (4,6).

Recent in vivo studies have confirmed that kisspeptin neurons in the arcuate nucleus of the hypothalamus are responsible for the periodic release of gonadotrophin-releasing hormone (GnRH) and subsequent pulsatile LH secretion that drives reproductive function (7–11). Studies using GCaMP fibre photometry in conscious mice have demonstrated that the arcuate kisspeptin neuronal population exhibits episodes of increased intracellular calcium levels coincident with, and immediately preceding, each pulse of LH secretion in intact and gonadectomised male and female mice (7–9). Miniscope investigations showed that individual kisspeptin neurons within the arcuate population act in a coordinated, synchronised, and episodic manner (10,11). Loss of pulsatile LH secretion during lactation and consequent lactational infertility may be caused by the loss of kisspeptin-mediated stimulation of GnRH secretion (12–18). Kisspeptin expression is markedly suppressed in lactation (12,16) and even when exogenously stimulated, kisspeptin neurons are unable to activate GnRH neurons during lactation, likely due to a lack of kisspeptin synthesis (13).

It is well established that hyperprolactinemia causes infertility, and thus, the elevated prolactin present in lactation seems a likely candidate to be involved in suppressing fertility during lactation. Prolactin administration acutely suppresses LH secretion (19), and chronic exposure to elevated prolactin reduces Kiss1 mRNA expression in the arcuate nucleus (17,20,21). In lactating mice, suppressing endogenous prolactin secretion shortens the period of infertility (22), suggesting that prolactin is important for maintaining the suppression of pulsatile LH secretion during lactation. Such a role for prolactin is controversial, however (4,6,23–27), with studies in a number of species suggesting that the neural stimulation of suckling may be more important than prolactin in maintaining lactational infertility (28,29). However, it has previously been difficult to disentangle the specific role of prolactin, as suckling, prolactin, and milk production are so tightly linked that manipulating one ultimately impacts the others, making it difficult to determine the contribution of any one element. Here, using a conditional deletion strategy, we have blocked prolactin action in the brain leaving suckling, lactation, and maternal behaviour intact. Using GCaMP fibre photometry techniques, we have also documented arcuate kisspeptin neuron activity across pregnancy and lactation transitions in the same mice and established that prolactin directly acts on these neurons to suppress fertility in lactation.

Methods

Animals

All experiments were performed using adult female mice on a C57BL/6J background (8-20 weeks of age unless otherwise stated). Mice were housed under controlled temperature (22 °C ± 2°C) and lighting (12-hour light/12-hour dark schedule, with lights on at 0600 hours) with ad libitum access to food and water (Teklad Global 18% Protein Rodent Diet 2918; Envigo, Huntingdon, United Kingdom). Daily body weight was recorded and daily vaginal cytology was used to monitor the estrous cycle stage. All experiments were carried out with approval from the University of Otago Animal Welfare and Ethics Committee.

To establish pregnancies, mice were mated with male wild-type C57BL/6J mice (presence of sperm plug = day 1 pregnancy). Male mice were then removed once a sperm plug was seen, and no male mice were present at parturition. The first day a litter was seen was counted as day 1 of lactation and maternal mice were left undisturbed till day 3 of lactation, when vaginal monitoring would resume and litter size was normalised to 6 pups per animal, unless otherwise stated. The intention of the normalisation was to reduce variability in the degree of suckling stimulus each dam would receive.

Towards the end of each recording period, bilateral ovariectomy (OVX) was performed under isoflurane anaesthesia with pre- and post-operative Carprofen (5mg/kg, s.c.) administered for pain relief. This enabled us to examine post-OVX changes in activity of kisspeptin neurons and expression of kisspeptin in the arcuate nucleus.

Pulsatile hormone measurement

To monitor pulsatile secretion of LH, serial tail tip blood sampling and measurement of LH by ELISA was undertaken as reported previously (19,30,31). As novel exposure and restraint stress has been shown to suppress pulsatile LH secretion (32), all mice were habituated to the tail tip blood sampling procedure by picking up the mouse in a gentle restraint device (soft cardboard tube) or hand and lightly massaging the tail for approximately 5 minutes per day for at least 3 weeks prior to experimentation (33). Sequential whole blood samples (4μl) were collected in 6 minute intervals for 3 hours between 0900 and 1200 hours, unless otherwise stated. Samples were immediately diluted in 48ul 0.01M PBS/0.05% Tween 20, and frozen on dry ice before being stored at -20°C for subsequent LH measurement.

RNAscope

Mice were transcardially perfused with a micro-perfusion pump with 2% paraformaldehyde (PFA). Brain sections (14μm-thick) were prepared, thaw mounted onto superfrost-plus microscope slides and then stored at -80°C. RNAscope in-situ hybridization was performed using the RNAscope 2.5 High definition Duplex Detection kit – chromogenic (Advanced Cell Diagnostics, Hayward, CA) largely in accordance with manufacturer’s instruction. The channel 1 Prlr probe was custom designed to pick up only the long form of the prolactin receptor. It was designed to transcript NM_011169.5 with a target sequence spanning nucleotides 1107-2147 (Ref: 588621; Advanced Cell Diagnostics, Hayward, CA). The channel 2 Kiss1 probe was custom designed to transcript NM_178260.3 with a target sequence spanning nucleotides 5 to 485 (Ref: 500141-C2; Advanced Cell Diagnostics, Hayward, CA). Sections were thawed at 55°C, postfixed for 3 minutes in 2% PFA, washed in 0.01M PBS for 5 minutes, and endogenous peroxidases were blocked with a hydrogen peroxidase solution for 10 minutes. Tissue was washed in distilled water (3x 2 minutes), then briefly immersed in 100% ethanol, air dried for 5 minutes, and a hydrophobic barrier was applied. Tissue was permeabilized with RNAscope protease plus for 30 minutes at 40°C. Sections were washed (2 x 2 minutes) and were hybridized with the Prlr and Kiss1 probes (1:300 dilution, Prlr:Kiss1) or negative control probe (Cat#320751; Advanced Cell Diagnostics, Hayward, CA) at 40°C for 2 hours. Amplification (Amp 1-6) was performed in accordance with the manufacturer’s instructions. Sections were then hybridized with a Fast-RED (1:60, Fast-RED B:Fast-RED A) for 10 minutes at room temperature, before undergoing further amplification steps (Amp 7-10) in accordance to manufacturer’s instructions. The final positive hybridization was detected by incubation with the secondary detection reagents (1:50, Fast-GREEN B:Fast-GREEN) for 10 minutes at room temperature. Sections were washed, counterstained with haematoxylin (25% Gills), dried at 60°C for 20 minutes, and cover-slipped with VectaMount (Vector laboratories, H-5000) before imaging using an Olympus BX51 light microscope and Olympus UPlanSApo 10/20x lenses.

Quantification of the proportion of kisspeptin neurons co-expressing Prlr mRNA was undertaken in FIJI software (National Institute of Health, Bethesda, Maryland, USA) following image acquisition. Positive hybridization for Kiss1 and Prlr was determined based on scoring guidelines included in the manufacturer’s protocol (Advanced Cell Diagnostics). The total number of Kiss1-expressing cells and the total number of these that showed Prlr mRNA expression were counted.

Effect of neuron-specific deletion of the prolactin receptor gene on the maintenance of lactational infertility

To investigate whether prolactin action in the brain is required for lactational infertility, neuron-specific Prlr knockout mice (Prlrlox/lox/Camk2aCre) and their respective Cre-negative controls (Prlrlox/lox) were generated (as previously described in (34)). We have previously shown that while Prlrlox/lox/Camk2aCre mice do not have a complete Prlr deletion in the forebrain. There are areas of extensive deletion (as measured by reduced prolactin-induced pSTAT5), such as the arcuate nucleus and ventromedial nucleus of the hypothalamus, and there are also areas where Prlr is reduced by about 50% such as the medial pre-optic area (34,35). In our experience with these animals, this is sufficient to retain normal maternal behaviour in most animals (our unpublished data), and this was the case for the animals used in the present study. RNAscope in-situ hybridization was done to confirm the degree of knockout in intact, diestrous, nulligravid, non-lactating (henceforth termed NL) mice and 14 day post-OVX mice (Prlrlox/lox intact n = 6, Prlrlox/lox OVX n = 5, Prlrlox/lox/Camk2aCre intact n = 5, Prlrlox/lox/Camk2aCre OVX n = 5; all aged 8-16 weeks). Both intact and OVX mice were included as kisspeptin cell bodies are only visible in the rostral periventricular region of the third ventricle (RP3V) of intact mice and in the arcuate nucleus of OVX mice, due to estradiol regulation (36). Prlrlox/lox/Camk2aCre mice showed a significant decrease in the percentage of Kiss1-expressing cells co-expressing Prlr compared to controls in both the RP3V (p = <0.0001) and arcuate nucleus (p = 0.0009) (unpaired two-tailed t tests, Supplementary Figure 1A-D).

Prlrlox/lox/Camk2aCre mice are hyperprolactinaemic due to impaired negative feedback of prolactin on hypothalamic dopamine neurons (34) and therefore show disrupted estrous cycles (showing recurrent pseudopregnancy-like cycles with long periods of diestrus of approximately 14 days between estrous stages). However, these mice are able to become pregnant and have normal pregnancies. All Prlrlox/lox/Camk2aCre mice were given a 250μl subcutaneous injection of bromocriptine (5mg/kg, 5% ethanol/saline; Tocris Bioscience Cat#0427) prior to being mated. This treatment was designed to reinstate an estrous cycle in Prlrlox/lox/Camk2aCre mice. Bromocriptine is an agonist for the type 2 dopamine receptor and inhibits prolactin secretion from the pituitary gland (37), thereby terminating the pseudopregnancy-like state and bringing the mice into proestrus the following day. Following treatment, all mice were then housed with a stud male.

For Prlrlox/lox/Camk2aCre mice, estrous cycles were monitored from day 3 of lactation until the first day of diestrus following a day of estrus (proestrus and estrus had to be observed prior to transcardial perfusion on the first day of diestrus). Brains were collected following transcardial perfusion for assessment of kisspeptin immunoreactivity. For every lactating Prlrlox/lox/Camk2aCre mouse (n = 8), the brain of a Prlrlox/lox control mouse (n = 8) of the equivalent day (±1) of lactation was also collected. A group of NL mice of both genotypes (n = 5-6) was also perfused for immunohistochemistry on diestrus.

To evaluate pulsatile LH secretion in early lactation (prior to the return of estrous cycles) and to determine whether progesterone played any role in regulating pulsatile LH secretion in lactation, additional groups of lactating Prlrlox/lox/Camk2aCre and Prlrlox/lox control mice were generated. These mice were treated with either the progesterone receptor antagonist, mifepristone (4mg/kg in sesame oil, s.c.; AK Scientific Inc Cat#J10622), or vehicle (n = 7-8 per group) on the morning of day 4 of lactation and day 5 of lactation. This dose was selected as it was found to be sufficient to cause termination of pregnancy in wild-type C57BL/6J mice (p = 0.0072, chi-squared test, Supplementary Figure 2A; pilot study, n = 6 both groups). Neither vehicle nor mifepristone treatment had an effect on litter weight gain (interaction of time x genotype & treatment p = 0.5322, two-way repeated measures ANOVA, Supplementary Figure 2B). Mice underwent blood sampling for LH for 3 hours on day 5 of lactation, 30 minutes after treatment.

Measurement of LH concentrations

An established sandwich ELISA method was used to determine LH concentration in diluted whole blood samples collected from mice (30,31). Briefly, a 96-well high plate was incubated with bovine monoclonal antibody (LHβ518b7, 1:1000 in 1xPBS; Dr. L. Sibley, UC Davis, CA, USA) for 16 h at 4°C. Following incubation of standards, controls and experimental samples for 2 hours, plates were incubated in rabbit polyclonal LH antibody (AFP240580Rb; 1:10,000; National Hormone and Pituitary Program, NIH) for 90 minutes, followed by incubation with polyclonal goat anti-rabbit IgG/HRP antibody (1:1000; DAKO Cytomation) for 90 minutes. Finally, plates were incubated in OPD (o-phenylenediamine capsules; Sigma-Aldrich Cat#P7288) for 30 minutes. A standard curve for the detection of LH concentration was generated using serial dilutions of mouse LH-reference preparation peptide (National Hormone and Pituitary Program, NIH). LH levels were read using a standard absorbance plate reader (SpectraMax ABS Plus; Molecular Devices) at 490nm and 630nm wavelengths. The assay had a sensitivity of 0.04ng/ml to 4ng/ml, with an intra-assay coefficient of variation of 4.40% and an inter-assay coefficient of variation of 8.29%.

PULSAR Otago was used to define LH pulses (38). Parameters used; Smoothing 0.7, Peak split 2.5, Level of detection 0.04, Amplitude distance 3, Assay variability 0, 2.5, 3.3, G(1)=3.5, G(2)=2.6, G(3)=1.9, G(4)=1.5, G(6)=1.2. Mean LH levels were calculated by averaging all LH levels collected during the experiment. Individual LH profiles from mice are shown in Supplementary Figure 3.

Assessment of kisspeptin expression

Perfusion and fixation of tissue

Mice were anaesthetised with sodium pentobarbital (15mg/mL, i.p.) and transcardially perfused with 4% PFA. Brains were removed, postfixed in the same solution, and cryoprotected overnight in 30% sucrose before being frozen at -80°C. Two sets of 30μm thick coronal brain sections were cut using a sliding microtome, from Bregma 1.10mm to - 2.80mm. Brain sections were kept in cryoprotectant solution (pH = 7.6) at -20°C until immunohistochemistry was performed.

Immunohistochemistry

Immunohistochemistry for kisspeptin in the RP3V and arcuate nucleus was performed as previously described (39). Briefly, sections were incubated in polyclonal rabbit anti-kisspeptin primary antibody (AC 566, 1:10,000; gift from A. Caraty, Institut National de la Recherche Agronomique, Paris, France) for 48 hours at 4°C. Sections were then incubated with biotinylated goat anti-rabbit IgG (1:200, Vector biolabs, Peterborough, GK) for 90 min at room temperature, followed by incubation in an avidin-biotin complex (Elite vectastain ABC kit, Vector laboratories). The bound antibody-peroxidase complex was visualised using a nickel-enhanced diaminobenzidine (DAB) reaction, to form a black cytoplasmic precipitate.

Brain sections were imaged as described in RNAscope section above. Quantification of kisspeptin neurons in the RP3V, was undertaken by manually counting all labelled neurons present in all three subdivisions: the anteroventral periventricular nucleus (AVPV); rostral preoptic periventricular nucleus (rPVpo); and caudal preoptic periventricular nucleus (cPVpo). Two sections per subdivision per mouse were averaged for each animal. As kisspeptin cell bodies in the arcuate nucleus were not easily observed, as previously reported (40), kisspeptin fibre immunoreactivity was imaged using a Gryphax NAOS colour camera (Jenoptik) and evaluated using FIJI software and the voxel counter function (National Institutes of Health). Kisspeptin fibre density was measured in the arcuate nucleus across the three subdivisions; rostral arcuate (rARC), middle arcuate (mARC), and caudal arcuate (cARC). Two sections of each subdivision per animal were counted and then averaged across each animal to get total number and reported as total amount of voxels per ROI (voxel fraction).

Characterization of arcuate kisspeptin neuronal activity using GCaMP fibre photometry

Stereotaxic surgery and AAV injections

Adult Kiss1Cre or Prlrlox/lox/Kiss1Cre mice (2-3 months old) were anaesthetised with 2% Isoflurane, given local Lidocaine (4mg/kg, s.c.) and Carprofen (5mg/kg, s.c.) and placed in a stereotaxic apparatus. A custom-made unilateral Hamilton syringe apparatus holding one Hamilton syringe was used to perform unilateral injections into the arcuate nucleus. The needles were lowered into place (-0.14mm A/P, +0.04mm M/L, -0.56mm DV) over 2 minutes and left in situ for 3 minutes before injection was made. 1μl AAV9-CAG-FLEX-GCaMP6s-WPRE-SV40 (1.3x10-13 GC/ml, University of Pennsylvania Vector Core, Philadelphia, PA, USA) was injected into the arcuate nucleus at a rate of ∼100nl/min with the needles left in situ for 3 minutes prior to being withdrawn over a period of 6 minutes. This was followed by implantation of a unilateral indwelling optical fibre (400 μm diameter, 6.5 mm long, 0.48 numerical aperture (NA), Doric Lenses, Canada, product code: MFC_400/430-0.48_6.5mm_SM3*_FLT) at the same coordinates. Carprofen (5mg/kg body weight, s.c.) was administered for post-operative pain relief. After surgery, mice received daily handling and habituation to the fibre photometry recording conditions over 4-6 weeks before experimentation began.

GCaMP fibre photometry

Photometry was performed as reported previously (9). Fluorescence signals were acquired using a custom-built fibre photometry system made primarily from Doric components. Violet (405nm) and blue (490nm) fibre-coupled LEDs were sinusoidally modulated at 531 and 211 Hz, respectively, and focused into a 400μm, 0.48 numerical aperture fibre optic patch cord connected to the mouse. Emitted fluorescence was collected by the same fibre and focused onto a femtowatt photoreceiver (2151, Newport). The two GCaMP6s emission signals were collected at 10 Hz in a scheduled 5s on/15s off mode by demodulating the 405nm (non-calcium dependent) and 490nm (calcium dependent) signals. The power output at the tip of the fibre was set at 50μW. Fluorescent signals were acquired using a custom software acquisition system (Tussock Innovation, Dunedin, New Zealand) and analysed using custom templates created by Dr Joon Kim (University of Otago, Dunedin, New Zealand) based on mathematics and calculations similar to those previously described (41,42). Briefly, the fluorescent signal obtained after stimulation with 405nm light was used to correct for movement artefacts as follows: first, the 405nm signal was filtered using a savitzky-golay filter and fitted to the 490nm signal using least linear square regression. The fitted 405nm signal was then subtracted and divided from the 490nm signal to obtain the movement and bleaching corrected signal. The output of these templates is 490-adjusted405/adjusted405, which was multiplied to get the final ΔF/F as a percentage increase. All photometry data is reported as ΔF/F(%).

All recordings were obtained from freely behaving mice between 0800 hours and 1200 hours (apart from 24 hours post weaning recording (0900 hours to 1700 hours), and day 18/19 pregnancy recording (1800 hours to 0800 hours the following day)). Synchronized events (SEs) were defined as when ΔF/F exceeds 3 standard deviations (SD) above the trace mean. Manual SE shape analysis was performed in addition to standard deviation method for certain datasets where necessary. SEs were counted manually to determine frequency of SEs per 60 minutes. The between-animal variability in total signal means that changes in SE amplitude can only be reported as relative changes within an animal. Relative SE amplitude was calculated by using normalised ΔF/F data and then subtracting the peak of an SE from the nearest nadir to the rise of the SE and averaging that for the number of SEs in a recording. To obtain normalised ΔF/F, three pre-pregnancy datasets from each mouse were used to find the average maximum ΔF/F for that mouse. All datasets were then divided by that normalisation value to get normalised ΔF/F for each trace for each individual mouse. In these long-term longitudinal studies, some data points are missing due to issues with recordings (e.g. dirty fibre optic cannula or system battery issues), meaning that occasionally fewer mice are used for analysis at various time points. During experimentation, no mice lost cannulas, but a gradual reduction in amplitude of signal was observed over time.

Following experimentation, brains were collected in a similar way to that described above. Postmortem analysis of cannula placement and GCaMP6 transfection in each animal was not done, because with experience, the quality and characteristics of the SE recordings (including corresponding pulsatile LH secretion) in each animal was indicative of successful cannula placement and transfection of the GCaMP. In a preliminary trial the green fluorescent protein (GFP) attached to the GCAMP was restricted to the caudal arcuate nucleus (n = 3 mice, preliminary trial), with little evidence of spread into the rostral or middle arcuate. GFP was seen exclusively ipsilaterally, with no GFP observed on the contralateral side. Cannula placement in these trial animals was in the caudal arcuate.

Monitoring the activity of arcuate kisspeptin neurons across different reproductive stages in the same mice

Adult Kiss1Cre mice were 8-10 weeks of age at the beginning of experiments, and up to 12 months old by the time of final recording (n = 8 during pregnancy; n = 6 during lactation, 2 mice were euthanised due to dystocia). Monitoring of vaginal cytology and weight was continuous from 1 week pre-surgery till day 19 of pregnancy and resumed on day 3 of lactation until the end of experiment. All handling stopped on day 19 of pregnancy to avoid potential compromise of parturition and onset of maternal behaviour.

To investigate the activity of the arcuate kisspeptin population across different reproductive states in the same animal, the following recording protocol was followed for all Kiss1Cre mice, unless otherwise stated; diestrus NL, day 4 of pregnancy, day 14 of pregnancy, day 18/19 of pregnancy (overnight), day 7 of lactation, day 14 of lactation, day 18 of lactation, 24 hours after weaning, first diestrus after estrous cycles begin following weaning, and 10 days post-OVX. In addition, blood sample collection for paired LH measurement was done in diestrus NL state, on day 14 of pregnancy (in 4 mice, maximum of 6 samples were collected around small “peaks” in baseline), day 7 of lactation, and day 14 of lactation. During sampling, blood samples were taken at the usual 6 minute intervals previously described, however if a SE was observed and the next scheduled sampling was more than 3 minutes away, an extra blood sample was taken 2 minutes after the SE. Following this additional sampling, the next scheduled blood sample would be taken at its original time and then every 6 minutes after, till another SE was seen. Blood sampling was not carried out at additional time points as blood sampling was undertaken at least a week apart, and stress from repeated sampling was attempted to be kept at a minimum e.g., not blood sampling around the time of birth. Fibre photometry recordings were usually between 2-4 hours in length. The only longer recordings were undertaken on day 18/19 of pregnancy (14 hours) and 24 hours after weaning (8 hours). These particular recording sessions were extended to determine whether there were any longer-term changes occurring in the activity of the arcuate kisspeptin population in the lead up to parturition or following weaning of pups (states closely followed by postpartum estrus and resumption of normal estrous cycles, respectively).

Effect of arcuate kisspeptin neuron-specific deletion of the prolactin receptor gene on the maintenance of lactational infertility and the activity of kisspeptin neurons during lactation

Kisspeptin-specific prolactin-receptor knockout mice (Prlrlox/lox/Kiss1Cre) and their respective Cre-negative controls (Prlrlox/lox) were generated as previously described (19). Prlrlox/lox/Kiss1Cre (n = 27) and Prlrlox/lox control (n = 30) dams underwent estrous cycle monitoring from day 3 of lactation onwards to determine whether mice showed an early resumption of estrous cycles. RNAscope in-situ hybridization (as described above) was done to confirm knockdown in a subgroup of these mice, with Prlrlox/lox/Kiss1Cre mice showing a significant decrease in the percentage of Kiss1-expressing cells co-expressing Prlr compared to controls in the arcuate nucleus (p = <0.0001, unpaired two-tailed t test, Supplementary Figure 1E-G; Prlrlox/lox n = 6, Prlrlox/lox/Kiss1Cre n = 8).

To determine whether the deletion of the prolactin receptor from arcuate kisspeptin neurons led to early reactivation of these neurons during lactation, adult Prlrlox/lox/Kiss1Cre mice (n = 5) and an additional Kiss1Cre control mouse (n = 1) (8 weeks old at the start of the experiment, and up to 14 months at final recording timepoint) were set up for fibre photometry, as described above. Monitoring of vaginal cytology and weight was continuous from 1 week pre-surgery till day 18 of pregnancy and resumed on day 3 of lactation. Recordings were undertaken in a similar timeline to that described above, however no pregnancy recordings were done, and in early lactation recordings were performed every 2 days from day 3 to day 9 of lactation, before following the same protocol as described. No blood samples were taken over the lactation period in these mice. As described previously, recordings were kept between 2-4 hours, apart from the 24 hours after weaning recording (8 hours).

Statistical analysis

Data are presented as mean ± SEM and all statistical analysis was performed with PRISM software 10 (GraphPad Software, San Diego, CA, USA) with a p value of < 0.05 considered as statistically significant. Individual symbols in graphs represent individual mice. Differences in kisspeptin cell number or fibre density was assessed using two-way ANOVAs with Tukey’s multiple comparisons tests or t tests, with both analyses using combined averages of each animal (averaged number of cells or fibre density across the three subdivisions of each nucleus to get total number reported). Resumption of estrous cycles was analysed using Log-rank (Mantel-Cox) test chi-squared test. LH pulse frequency data and mean LH data was analysed using two-way ANOVAs with Tukey’s multiple comparisons tests. SE frequency and amplitude throughout reproductive cycles was analysed using mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests where appropriate and day 18/19 of pregnancy data was analysed using t tests. Correlation between SE occurrence and LH pulses was assessed using chi-squared test. All fibre photometry data used for quantitative analysis and comparison were from the first pregnancy and lactation. A full list of probability values, inferential statistics, and degrees of freedom for all data can be found in Supplementary Table 1.

Results

Prolactin action on forebrain neurons is necessary to maintain lactational infertility

Lactation has previously been associated with a marked decrease in Kiss1 mRNA levels in both RP3V and arcuate nucleus populations (13). To determine whether prolactin was involved in the maintenance of lactational anestrus, the Prlr gene was knocked out of Camk2a expressing neurons (most forebrain neurons) of female mice. Control Prlrlox/lox mice showed a marked reduction in kisspeptin cell body immunoreactivity in the RP3V of lactating compared to NL mice (p = 0.0100, Post hoc Tukey’s multiple comparisons test, Figure 1A, C). In contrast, in Prlrlox/lox/Camk2aCre mice the lactation-induced suppression of kisspeptin cell bodies in the RP3V was absent (p = 0.6409, Post hoc Tukey’s multiple comparisons test; interaction between reproductive state and genotype p = 0.0034, two-way ANOVA, Figure 1A, C). Fibre density in the arcuate nucleus was significantly increased in lactating Prlrlox/lox/Camk2aCre mice compared to lactating Prlrlox/lox controls (p = 0.0020, unpaired two tailed t test, measured as percentage voxels within the region of interest, Figure 1B, D).

Prlrlox/lox/Camk2aCre mice do not undergo the normal period of lactational infertility and the lactation-induced suppression of kisspeptin immunoreactivity is absent.

(A) Kisspeptin immunoreactivity shown in representative photomicrographs from diestrus, nulligravid, non-lactating (NL; left) and lactating (right) Prlrlox/lox control and Prlrlox/lox/Camk2aCre mice (from anteroventral periventricular nucleus (AVPV) region of RP3V). (B) Representative photomicrographs showing mid arcuate nucleus (mARC) of lactating Prlrlox/lox (top) and lactating Prlrlox/lox/Camk2aCre mice (bottom). (C) Total kisspeptin cell number for the RP3V (NL Prlrlox/lox (n = 6) versus lactating Prlrlox/lox control (n = 8) p = 0.0100, NL Prlrlox/lox/Camk2aCre (n = 5) versus lactating Prlrlox/lox/Camk2aCre (n = 8) p = 0.6409). Two-way ANOVA followed by Tukey’s multiple comparisons test. (D) Quantification of kisspeptin fibre density in the arcuate nucleus (Fiji software, measured in percentage voxels per region of interest), showing total kisspeptin fibre density in the arcuate nucleus (lactating Prlrlox/lox control n = 8, lactating Prlrlox/lox/Camk2aCre n = 7, p = 0.0020, unpaired two-tailed t test). (E) Lactating Prlrlox/lox/Camk2aCre mice (blue, n = 8) resume estrous cycles significantly earlier (100% within 6-10 days of lactation) than lactating Prlrlox/lox controls (grey, n = 10) (p = <0.0001, Log-rank (Mantel-Cox) test). Scale bar image and insert = 50μm. Values are shown as mean ± SEM.

Estrous cycles during lactation were significantly altered by deletion of Prlr in the forebrain, with all Prlrlox/lox/Camk2aCre mice showing a return to estrus between day 6 and day 10 of lactation (Figure 1E), while, as normal, estrus did not occur until day 20 (after weaning) in control Prlrlox/lox mice (p = <0.0001, log-rank (Mantel-Cox) test, Figure 1E). No differences in litter weight gain from day 3 to day 8 of lactation were observed in either group (p = 0.3282, mixed analysis test, Supplementary Figure 4), indicating that the suckling stimulus that mice received and lactating itself was maintained in the absence of Prlr expression in Camk2a expressing neurons. Collectively, these data show that prolactin action in the brain is absolutely required for the lactation-induced suppression of kisspeptin expression and to maintain lactational infertility in mice.

In a separate cohort of Prlrlox/lox/Camk2aCre mice, pulsatile LH secretion was monitored in early lactation, prior to the return of estrous cycles. To rule out a potential role for progesterone in suppressing fertility during lactation (9,43,44), the progesterone receptor antagonist mifepristone (RU486), was administered to these mice in early lactation. Vehicle-treated Prlrlox/lox control mice showed the expected near complete absence of pulsatile LH secretion during lactation (Figure 2A, B). In contrast, nearly all Prlrlox/lox/Camk2aCre mice showed a lack of the normal lactation-induced suppression of pulsatile LH secretion demonstrated by a significant increase in frequency of LH pulses compared to controls (effect of genotype p = 0.0024, two-way ANOVA with Tukey’s multiple comparisons test, Figure 2A, B). There was no effect of mifepristone on the pattern of LH secretion, suggesting that progesterone action is not required for the suppression of LH secretion (LH pulse frequency (interaction genotype x treatment p = 0.2807; treatment p = 0.8588; Figure 2B); mean LH levels (interaction genotype x treatment p = 0.8697; treatment p = 0.8586; Figure 2C); two-way ANOVA with Tukey’s multiple comparisons test). These data indicate that prolactin is the primary signal responsible for the suppression of pulsatile LH secretion during lactation in mice.

Prolactin action in the brain during lactation is necessary for the suppression of pulsatile LH secretion.

Examples of pulsatile LH levels in the blood from lactating Prlrlox/lox control and lactating Prlrlox/lox/Camk2aCre mice that have either been treated with vehicle (sesame oil, s.c., grey, veh) or 4mg/kg mifepristone (in sesame oil, s.c., blue, mif). Graphs show LH pulse frequency (B; interaction p = 0.2807, genotype p = 0.0024, state p = 0.8558), and mean LH levels (C; interaction p = 0.8697, genotype p = 0.0031, state p = 0.8586). Lactating vehicle-treated Prlrlox/lox (n = 8), lactating mifepristone-treated Prlrlox/lox (n = 8), lactating vehicle-treated Prlrlox/lox/Camk2aCre (n = 8), lactating mifepristone-treated Prlrlox/lox/Camk2aCre (n = 7). Asterisks indicate LH pulse peaks as detected by PULSAR Otago analysis. Two-way ANOVA followed by Tukey’s multiple comparisons test. Values are shown as mean ± SEM.

Episodic activity of arcuate kisspeptin neurons is suppressed during pregnancy and most of lactation

To directly assess the role of prolactin in regulating kisspeptin neuron activity during lactation, GCaMP fibre photometry was undertaken to monitor real-time activity of arcuate kisspeptin neurons in freely behaving mice. We first undertook a longitudinal assessment of changes in kisspeptin neuronal activity by tracking individual Kiss1Cre mice throughout pregnancy, lactation, and following weaning (Figure 3).

Longitudinal recordings of arcuate kisspeptin neuron GCaMP population activity throughout different reproductive states in the same Kiss1Cre mice.

Representative neuronal activity from three Kiss1Cre mice throughout the NL (diestrus, nulligravid, non-lactating), pregnant, lactating, and post-weaning states. The time points monitored in order were: diestrus NL, day 4 pregnancy, day 14 pregnancy, day 18/19 pregnancy (overnight), day 7 lactation, day 14 lactation, day 18 lactation, 24 hours after weaning (day 22 postpartum), return to normal cycling following weaning (return to estrous cycles), and 10 days post ovariectomy (OVX). Asterisks indicate SEs. Note: change in y axes scale on all 3 OVX timepoints and mouse shown in dataset (B) from day 4 of pregnancy onwards.

Initially, GCaMP fibre photometry recordings were collected in the NL diestrous state, both with and without serial blood sampling to measure LH concentrations. As can be seen in Figure 4A, arcuate kisspeptin photometry recordings were characterized by discrete SEs of elevated intracellular calcium (indicative of synchronous activity of the kisspeptin population), with each SE consistently correlating to a single pulse of LH secretion in the minutes following (p = <0.0001, chi-squared test). These were observed to be at a similar rate to that described previously in diestrous mice using GCaMP photometry in a different Kiss1 mouse line (9), with periodic SEs occurring about once per hour (1.250±0.250/hr; Figure 3 & 4B).

Synchronised Ca2+ events (SE) are consistently correlated with pulsatile LH secretion across different reproductive states in Kiss1Cre mice.

(A) When fibre photometry was paired with serial blood sampling for pulsatile LH secretion, the relationship between SEs and LH pulses was examined. Each time an SE was seen during a recording with blood sampling, pulsatile secretion of LH was also observed, with 100% correlation (p = <0.0001, chi-squared test; 73 out of 73 SEs observed led to an LH pulse). Representative examples of paired photometry and blood sampling are shown from the NL (diestrus, nulligravid, non-lactating) state, from day 7 lactation, and from day 14 lactation. (B) Quantitative analysis of SE frequency per hour across different reproductive states in Kiss1Cre mice (p = 0.0012, mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests). (C) Quantitative analysis of relative SE amplitude of normalised ΔF/F across different reproductive states (p = 0.0118, mixed effect analysis (fixed type III) with Tukey’s multiple comparisons tests). Black asterisks indicate SEs, red asterisks indicate LH pulse peaks as detected by PULSAR Otago analysis. Values shown as mean ± SEM.

The activity of the arcuate kisspeptin population in Kiss1Cre mice dynamically changed depending on the reproductive state of the mouse (p = 0.0012, mixed-effect analysis, Figure 3 & 4B). On day 4 of pregnancy, SE frequency had markedly decreased (0.297±0.136/hr; Figure 3 & 4B), indicating an early reduction in activity of arcuate kisspeptin neurons during pregnancy. By day 14 of pregnancy, no SEs were seen (0±0/hr; Figure 3 & 4B) and this was confirmed by a lack of pulsatile LH secretion (Supplementary Figure 5). In late pregnancy (day 18), neuronal activity was monitored for 14 hours, during which time low amplitude, SEs were unexpectedly observed at similar frequencies to NL levels (2.043±0.940/hr; Figure 3 & 4B-C). This unusual pattern of activity is illustrated in more detail in Figure 5 where, alongside the resurgence of low amplitude SEs, a marked increase in baseline activity was observed relative to recordings at other reproductive stages. This activity was reminiscent of the miniature SEs observed to be caused by activation of subgroups of arcuate kisspeptin cells using brain slice calcium imaging (11), but we are unable to resolve such SEs using the present methods. Since our aim was to continue longitudinal assessment of the arcuate kisspeptin population into lactation and weaning, mice were not disrupted by blood sampling immediately prior to parturition as we were concerned that this additional stressor might interfere with establishment of maternal behaviour. Hence, we are unable to report whether these low amplitude SEs and elevated baseline activity were associated with LH secretion.

Activity of arcuate kisspeptin neurons on day 18 and 19 of pregnancy in Kiss1Cre mice.

Fibre photometry recordings of mice on the evening of day 18 of pregnancy (1800 hours) to the morning of day 19 of pregnancy (0800 hours) show low amplitude SEs. 3-hour section blown up for ease of viewing. (C) No difference was seen between frequency of SEs (per 60 minutes) in the NL (diestrus, nulligravid, non-lactating) versus day 18/19 pregnancy (p = 0.4063, paired two-tailed t test), however a significant decrease in relative SE amplitude is seen (D; p = 0.0141, paired two-tailed t test). Asterisks indicate SEs. Dotted line in insert of (A) and (B) indicates 3 standard deviations (3SD). Grey shaded region = lights off.

Evaluation of activity of arcuate kisspeptin neurons during lactation showed complete suppression of activity on day 7 of lactation with a corresponding absence of pulsatile LH secretion (0±0/hr; Figure 3 & 4B). This lactation-induced suppression of activity was partially relieved by day 14 of lactation (0.583±0.083/hr; Figure 3 & 4B), with low frequency SEs again corresponding to low frequency pulsatile LH secretion. Further increases in SE frequency were seen on day 18 of lactation (0.850±0.100/hr; Figure 3 & 4B), including an increase in baseline activity, similar to that seen in late pregnancy, and by 24 hours after weaning (day 22 postpartum) the frequency of SEs had returned to close to non-pregnant levels (1.375±0.114/hr; Figure 3 & 4B). SE frequency remained unchanged on the day of first diestrus following a return to estrous cycles after weaning (1.533±0.226/hr; Figure 3 & 4B). Ten days post-OVX, clusters of high amplitude arcuate kisspeptin population activity was observed (4.778±0.222/hr; Figure 3 & 4B), consistent with previous reports following OVX in nulliparous mice (45). Collectively, these observations show extensive, dynamic variation in the activity of the arcuate kisspeptin neuronal population associated with pregnancy and lactation.

Mice with an arcuate kisspeptin neuron-specific deletion have premature reactivation of estrous cycles and neuronal activity in lactation

To determine whether the prolactin-induced suppression of estrous cycles and pulsatile LH secretion was specifically mediated by kisspeptin neurons, mice were generated with an arcuate-specific deletion of the Prlr from kisspeptin neurons (19). Similar to the data from Prlrlox/lox/Camk2aCre mice, there was early resumption of estrous cycles in Prlrlox/lox/Kiss1Cre mice during lactation (63% showing estrus by day 10 of lactation and 83% by day 19 lactation) compared to Prlrlox/lox controls (4% by day 19 lactation) (p = <0.0001, log-rank (Mantel-Cox) test, Figure 6A). No difference in litter weight gain during lactation was observed in either group indicating that suckling and/or lactation itself was not impaired (day 3-20 lactation weight gain, p = 0.6404, two-way ANOVA, Supplementary Figure 4H). In vivo GCaMP fibre photometry in Prlrlox/lox/Kiss1Cre mice showed early reactivation of the arcuate kisspeptin population by day 5 of lactation (Figure 6B). This was accompanied by a clear return to estrus within this early lactation window in four out of five mice. These data demonstrate, in mice, that prolactin action specifically on arcuate kisspeptin neurons is responsible for maintaining suppression of those neurons, and thereby fertility, during lactation.

Prlrlox/lox/Kiss1Cre mice do not undergo the normal period of lactational infertility and show early reactivation of arcuate kisspeptin neurons prior to estrus in lactation.

(A) Prlrlox/lox/Kiss1Cre mice resume estrous cycles significantly earlier (63% by day 10 of lactation, n = 32, blue) than Prlrlox/lox controls (4% by day 19 of lactation, n = 30, grey) (p = <0.0001, Log-rank (Mantel-Cox) test). (B) Representative fibre photometry traces from day 5 of lactation from either a Kiss1Cre control mouse or Prlrlox/lox/Kiss1Cre mice. Mice with Prlr knocked out of arcuate kisspeptin neurons (Prlrlox/lox/Kiss1Cre) show SEs early in lactation. Asterisks indicate SEs.

Discussion

We demonstrate here that prolactin action in arcuate kisspeptin neurons is necessary for the maintained suppression of fertility during lactation in mice. Neuron-specific Prlr deletion (Prlrlox/lox/Camk2aCre) resulted in premature return to estrus in early lactation, even in the presence of ongoing suckling stimulus and the full metabolic consequences of milk production. Accompanying the resumption of estrus was an absence of the normal lactation-induced reduction in kisspeptin immunoreactivity (12,16,20). Pulsatile LH secretion was also observed on day 5 of lactation prior to the premature estrus when it would normally have been completely absent (46–48). To evaluate the specific role of kisspeptin neurons in mediating the prolactin-induced suppression of fertility, we have comprehensively mapped the activity of arcuate kisspeptin neurons throughout a full reproductive cycle: pregnancy, lactation, and after weaning in individual animals. The data show an immediate suppression of activity of arcuate kisspeptin neuronal activity during pregnancy, and this is maintained throughout most of lactation, apart from a brief window of reactivation immediately prior to parturition. Deleting Prlr specifically from arcuate kisspeptin neurons (Prlrlox/lox/Kiss1Cre) prevented the suppression of activity in early lactation, resulting in premature induction of episodic activation of kisspeptin neurons, and early onset of estrus. Combined, these data provide direct evidence that prolactin action on kisspeptin neurons is necessary for lactation-induced infertility in mice.

It is now well established that arcuate kisspeptin neurons form the GnRH “pulse generator”, and hence drive pulsatile release of GnRH from the hypothalamus and consequent pulsatile secretion of LH from the pituitary that is required for fertility (49–51). This is the first study to monitor activity of the arcuate kisspeptin neurons across different reproductive states in the same animal, and the data largely match previously described patterns of LH secretion (47,52). The frequency and dynamics of the SEs changed dramatically, initially due to the pregnancy-induced changes in ovarian hormones. The abrupt decrease in arcuate kisspeptin activity in early pregnancy is likely caused by rising levels of progesterone, known to profoundly suppress activity of arcuate kisspeptin neurons and LH secretion (9,43,44). Progesterone is elevated throughout pregnancy, gradually increasing until luteolysis and progesterone withdrawal occurs in the lead up to parturition (53–56). Interestingly, we observed a transient reactivation of the arcuate kisspeptin neurons in the night between days 18 and 19 of pregnancy. This was characterised by frequent, low amplitude SEs, and increased baseline activity that may represent the intermittent synchroniszed activity of small subsets of arcuate kisspeptin neurons that have not yet transitioned to full synchroniszation of the whole population (11). It seems likely that this pattern of activity is associated with progesterone withdrawal in late pregnancy and may be important in stimulating follicular growth leading up to a postpartum ovulation (57,58).

In early lactation, episodic activity of arcuate kisspeptin neurons was absent, with sporadic low-amplitude activity returning around day 14 of lactation. There was another period of increased baseline activity in late lactation, similar to that seen in late pregnancy, potentially representing a signature of reactivation of synchroniszed activity of the arcuate kisspeptin neurons. This increase in arcuate kisspeptin population activity during late lactation mirrors the increase in LH levels that has been reported as lactation progresses (59). Overall patterns of SE activity rapidly returned to normal diestrus levels soon after weaning (59). However, in the absence of Prlr in arcuate kisspeptin neurons, SE activity re-appeared as early as day 5 of lactation, even in the presence of ongoing suckling. These data clearly show that prolactin action in arcuate kisspeptin neurons is necessary to sustain lactational infertility in mice. The observed disruption of lactational infertility in Prlrlox/lox/Kiss1Cre mice is particularly remarkable given that Prlr deletion is restricted to arcuate kisspeptin neurons in this model (19), and prolactin action on RP3V kisspeptin neurons (21,60) and on gonadotrophs in the pituitary gland (61–63) are unaffected.

The indispensable role for prolactin in mediating lactation-induced infertility in the mouse is surprising, given the consensus of much work in other species concluding that other factors may be more important (see (4,6,25,26,64)). This may reflect a level of redundancy amongst contributing factors across all species, including ovarian hormones, metabolic cues and neural inputs of suckling. Notably, the conditional deletion approach described here distinguishes prolactin action from the neurogenic effects of suckling without altering the process of lactation itself. Moreover, this approach avoids the potential confounding effects of using dopamine agonists to suppress prolactin (28,29), given that dopamine can directly inhibit GnRH neuronal activity (65).

While the effects of widespread neuronal deletion (Prlrlox/lox/Camk2aCre) on fertility were largely recapitulated by the arcuate kisspeptin-specific model (Prlrlox/lox/Kiss1Cre), it was apparent that the global deletion was more effective at inducing the return to estrus during lactation (in 100% of Prlrlox/lox/Camk2aCre mice by day 10), compared to the arcuate kisspeptin-specific model (63% by day 10, and 83% by day 19). This may be due to the absence of lactation-induced suppression of Kiss1 expression in RP3V kisspeptin neurons of Prlrlox/lox/Camk2aCre mice. Similarly, we cannot rule out the possibility that other populations of prolactin-sensitive neurons, such as GABA or dopamine neurons (60), may contribute to suppressing estrous cycles during lactation.

Nevertheless, our data collectively provide strong evidence that prolactin action on arcuate kisspeptin neurons is the primary factor mediating lactation-induced infertility in mice. Given that hyperprolactinemia induces infertility in many other mammalian species, including humans (21,22,66–75), it is possible that a conserved mechanism could be contributing to lactational infertility in these species.